{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}

### 2016-12-05 ident_manu.tex
first two pages of discussion

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^2$, Lisa L. Lowe$^3$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Email: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}}}
	  \\\\{\textit {\normalsize $^2$Dauphin Island Sea Lab, University of South Alabama}}
  \\{\textit {\normalsize Dauphin Island, AL 36528}}
	\\{\textit {\normalsize Phone: 251-861-2141 ext. 7567}}
	\\{\textit {\normalsize Email: \href{mailto:jlehrter@disl.org}{jlehrter@disl.org}}}
	\\\\{\textit {\normalsize $^3$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Thu Dec 1 15:17:07 2016 -0600
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

\begin{abstract}
\noindent Bio-geo-chemical models are useful tools in environmental sciences that can guide management and policy-making. Consequently, significant time and resources are spent developing these models in system-specific contexts. The optimization of model parameters to maximize precision, including transferability of these models to different systems, are fundamental concerns in the development and application of these tools. This study provides a context for understanding quantitative limitations of coupled hydrodynamic-ecological models by evaluating parameter sensitivity and identifability of a \ac{zerod} unit of a larger spatio-temporal model of hypoxia on the \acl{lcs} of Gulf of Mexico.  The analysis provides a contrast of numeric and ecological certainty in parameter subsets using a systematic framework to infer larger trends in dissolved oxygen dynamics over time, having implications for understanding factors that contribute to environmental conditions that are detrimental to aquatic resources.  In particular, we focus on issues of parameter identifiability using local sensitivity analyses to provide quantitative descriptions of numerical constraints on model precision.  We argue that quantitative and ecological certainty in model calibration are often at odds and practitioners must choose explicit model components to optimize given tradeoffs between the two. We further conclude that numerically optimal parameter sets for models of hypoxia are often small subsets of the complete parameter set because of redundancies in the unique effects of paramater perturbations on model output.  As a result, we demonstrate that use of a model for inference into ecological mechanisms of observed or predicted changes in hypoxic condition can be potentially misguided in the absence of quantitative descriptions of identifiability.  Although these concerns have been expressed in the literature, they are rarely explicitly addressed or included in model evaluations.  In addition to immediate implications for regional models, we provide a framework for describing the effects of parameter uncertainty and identifiability that can be applied to similar models to better inform environmental management.
\end{abstract}
\acresetall

@@

\noindent where $\gamma$ ranges from one to infinity for perfectly identifiable (orthogonal) or unidentifiable (perfectly collinear) results for parameters in a sensitivity matrix $S$.  The sensitivity functions were supplied as a matrix $\hat{S}$ with rows $i$ and columns $j$ (\cref{sijeqn}) that described deviations of predicted \ac{do} from the default parameter values.  The matrix $\hat{S}$ was first normalized by dividing by the square root of the summed residuals \citep{Omlin01,Soetaert10}.

The collinearity index $\gamma$ provides a measure of the linear dependence between sensitivity functions described above for subsets of parameters. Estimates of $\gamma$ greater than 10-15 suggest parameter sets are poorly identifiable \citep{Brun01,Omlin01}, meaning parameter values that maximize precision on a calibration dataset are inestimable by conventional optimization algorithms given similar effects of the selected parameters on \ac{do}. Greater sensitivity of a state variable to a subset of parameters does not always imply better identifiability if the individual effects are similar.  An intuitive interpretation of $\gamma$ is provided by \citet{Brun01} such that a change in a state variable caused by a change in one parameter can be offset by the fraction $1 - 1/\gamma$ by the remaining parameters.  That is, $\gamma = 10$ suggests the relative change in \ac{do} for an arbitrary parameter in the selected set can be compensated for by 90\% with changes in the other parameters.

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the 51 evaluated for the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter (e.g., practicioners may only be interested in optics parameters).  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability limits for parameter subsets, and the difficulty of choosing from many combinations.

@@

Local sensitivity analyses for the additional state variables (ammonium, \ac{chla}, irradiance, nitrate, \ac{pom}, \ac{dom}, and phosphorus) had similar results as \ac{do} with some exceptions (\cref{fig:sensalltile,tab:nh4sens,tab:chlsens,tab:irrsens,tab:no3sens,tab:om1sens,tab:om2sens,tab:po4sens}).  All additional variables were sensitive to the same parameters (38 of 51 evaluated), although average sensitivity differed between variables.  Average $L1$ ranged from $0.02$ for irradiance (\cref{tab:irrsens}) to $0.71$ for \ac{dom} (\cref{tab:om2sens}).  All average sensitivity values for the state variables were higher than the average for \ac{do} ($L1$ = $0.01$).  For each variable, $L1$ ranged from $2.24\times 10^{-6}$ (\textit{QminP}) to $8.49$ (\textit{mA}) for ammonium (\cref{tab:nh4sens}), $1.38\times 10^{-6}$ (\textit{QminP}) to $13.94$ (\textit{mA}) for \ac{chla} (\cref{tab:chlsens}), $1.92\times 10^{-7}$ (\textit{QminP}) to $0.13$ (\textit{ZKa}) for irradiance (\cref{tab:irrsens}), $6.67\times 10^{-7}$ (\textit{QminP}) to $8.49$ (\textit{umax}) for nitrate (\cref{tab:no3sens}), $6.41\times 10^{-5}$ (\textit{KNH4}) to $7.22$ (\textit{mA}) for \ac{pom} (\cref{tab:om1sens}),  $7.41\times 10^{-5}$ (\textit{KNH4}) to $14.25$ (\textit{mA}) for \ac{dom} (\cref{tab:om2sens}), and $8.21\times 10^{-7}$ (\textit{QminP}) to $1.47$ (\textit{ZKa}) for phosphate (\cref{tab:po4sens}).  For the parameter categories, ammonium was most sensitive to phytoplankton parameters (average $L1$ = $0.8$ across all parameters in the category), \ac{chla} to phytoplankton ($L1$ = $1.14$), irradiance to zooplankton ($L1$ = $0.03$), nitrate to zooplankton ($L1$ = $1.06$), \ac{pom} to temperature ($L1$ = $0.86$), \ac{dom} to temperature ($L1$ = $1.48$), and phosphate to zooplankton ($L1$ = $0.31$).  Finally, average sensititivy between parameter categories independent of the state variables ranged from $0.01$ for optics (average $L1$ across all variables) to $0.62$ for phytoplankton.

@@

The identifiability analyses suggested that many parameter subsets exceeded the thresholds of $\gamma = 10, 15$, providing further justification for using selection heuristics for parameter optimization.  Results for \ac{do} are provided first to demonstrate general concepts for the identifiability analyses, followed by an extension to the remaining state variables.  Parameter identifiability (as $\gamma$) for \ac{do} increased at different rates with increasing size of parameter subsets depending on the parameter category or the number of top parameters that were selected (\cref{fig:identplo}). By category, identifiability was lowest for all combinations of parameter subsets (from two to many parameters) in the phytoplankton ($60$\% less than $\gamma = 15$, $43$\% less than $\gamma = 10$) and zooplankton categories ($53.1$\% less than $\gamma = 15$, $40$\% less than $\gamma = 10$), whereas all combinations were identifiable for optics ($100$\% less than $\gamma = 15$, $100$\% less than $\gamma = 10$) and a majority identifiable for organic matter ($91.9$\% less than $\gamma = 15$, $76.5$\% less than $\gamma = 10$) combinations were identifiable. Identifiability for parameters in the temperature category was not evaluated because \ac{do} was sensitive to only one parameter (i.e., $\gamma = 1$). Parameter combinations for choosing from the top, top two, top three, and top four parameters in each category together had decreasing identifability with the increasing size of the selection pool (e.g., top one versus top four parameters, \cref{fig:identplo}).  The percentage of parameter subsets that were below the acceptable thresholds for identifability was $100$\% less than $\gamma = 15$ and $100$\% less than $\gamma = 10$ for the top parameters in each category, $90.6$\% and $80.7$\% for the top two, $80.7$\% and $70.9$\% for the top three, and $55.8$\% and $45.7$\% for the fop four.  Results for the remaining state variables had similar patterns in identifability with increasing size of parameter subsets and selection categories, although differences in identifiabiltiy between state variables was observed (\cref{fig:identploall}).  Most notably, nitrate was consistently the least identifiable variable for all selection heuristics (highest $\gamma$), whereas \ac{do} was most identifiable.

An alternative view of the results in \cref{fig:identplo} can be used to demonstrate the effects of parameter selection criteria and number of parameters in the selection pool on identifiability.  \Cref{fig:percthresh} shows the percentage of identifiable parameter sets for \ac{do} using the same selection criteria in \cref{fig:identplo}, i.e., selection of parameters only within parameter categories and selection of the top sensitive parameters regardless of category.  \Cref{fig:percthresh} is conceptually similar to \cref{fig:identplo}, with the added effect of a chosen $\gamma$ threshold on identifability.  Previous studies have provided general rules for $\gamma$ thresholds \citep{Brun01,Omlin01}, such that exact values beyond which model calibration fails could vary for the given model and optimization method.  The selected $\gamma$ thresholds are somewhat arbitrary, although the results show differences that have implications for parameter selection for model calibration.  In general, identifiability decreases with the addition of parameters, although the rate of decrease depends on the selected threshold for $\gamma$. More conservative values for $\gamma$ (e.g., $\gamma = 5$) were more sensitive to the number of parameters in a subset.  Notable differences in identifiability were also observed by selection criteria (within categories or top parameters only), which further supports results in \cref{fig:identplo,fig:identploall}.

An evaluation of the effects of individual parameters on $\gamma$ further suggested that some parameters have disproportionate effects on identifiability.  Based on $\gamma = 15$, \cref{fig:identplo} suggests that most parameter sets for organic matter were identifiable, regardless of how many parameters were selected (i.e., two through eight).  Because identifiability of the organic matter parameters was generally acceptable, identification of one or more redundant parameters that are inflating $\gamma$ values could provide useful information.  \Cref{fig:exclex} shows an alternative view of identifiability of \ac{do} with exclusion and inclusion of individual parameters in different sets for the organic matter category.  Collinearity increases with more parameters included in a subset, although the increase varies depending on which parameter was included or excluded from the set.  For example, inclusion of \textit{KNO3} in a parameter set almost always inflated $\gamma$.  All parameter subsets that did not include \textit{KNO3} were well below $\gamma = 15$, suggesting that removal of this parameter improves identifiability.  Interestingly, the inclusion of some parameters caused a reduction in $\gamma$, which contradicts the general rule that more parameters caused reduced identifiablity.  For example, parameter sets that included \textit{KGcdom} generally had lower $\gamma$ values relative to those that excluded the parameter.

@@

The above results demonstrated that each state variables differed in the magnitudes of sensitivity for each parameter and the number of identifiable subsets, where the latter varied by identifiability thresholds and parameter selection criteria.  These results further underscore the need for explicit selection heuristics for parameters to calibrate for improving model performance. Results for each of the three selection heuristics (blocked by parameter category, independent of category, all categories equally) applied to each state variable differed in the number of selected parameters and distribution of parameters within each category (\cref{tab:heurist1,tab:heurist2,tab:heurist3}).  In general, a corresponendence was observed between the number of parameters that were selected given the threhold of $\gamma = 15$ and relative identifiability between the state variables.  As noted above, nitrate was the least identifiable variable (\cref{fig:identploall}), whereas other variables (e.g., \ac{do}, irradiance) were more identifiable.  The constraints on relative identifiability between variables were demonstrated with the selection heuristics.  For example, heuristics for nitrate typically selected only one or two parameters that met the criteria as compared to more identifiable variables that included several parameters that were identifiable. Overall, the first selection heuristic demonstrated that the number of parameters chosen by parameter category differed independently of the state variables (\cref{tab:heurist1}). The number of selected parameters averaged across state variable in decreasing order was 4.25 parameters (phytoplankton), 3.5 (organic matter), 2.75 (optics), and 2.38 (zooplankton).  These rankings were independent of parameter sensitivity and identifiability in each category.  The second and third selection heuristics (\cref{tab:heurist2,tab:heurist3}) were similar, although more parameters were generally selected for the third heuristic.

\Cref{fig:heurist_stts} demonstrates parameter selection for all state variables following the second heuristic where parameters were chosen by decreasing sensitivity independent of parameter categories (exact values are shown in \cref{tab:heurist2}).  The y-axis on the plot shows the relative identifiability values with the addition of parameters from one to many on the x-axis (from left to right).  The second to last parameter for each variable is the last parameter selected within the potential threshold of $\gamma = 15$.  Interestingly, the last parameter shown for most of the state variables caused a relatively large increase in $\gamma$ that was disproportionate to the combined identifiability of the preceding parameter set. For most variables, the phytoplankton edibility vector for zooplankton, \textit{ediblevector(Z1)}, caused a dramatic increase in $\gamma$ with inclusion in the parameter set.  In addition to demonstrating the approach for selection with the second heuristic, \cref{fig:heurist_stts} provides an alternative approach to identifying parameters that are disporportionately redundant within a given parameter set.  This infomation is useful if additional parameter selection rules are developed ndependently of those proposed herein.

@@

\section{Discussion}
%' 
%' % back of envelop calcs for number of relevant parms, all state variables
%' <<echo = F>>=
%' 
%' @
%' We showed only small subsets are identifiable, similar conclusions have been described by citations in \citep{Wagener01}, p. 14 for models that follow traditional calibration schemes (e.g., objective function minimization).
%' 
%' Emphasize that parameters that have the greatest effect on collinearity are not those that have the highest sensitivity (contrast the identifiability by category vs identifiability by top parameters)  , also note that groups of parameters together can have large effects on collinearity, maybe some kind of bootstrap analysis could be done looking at doubletons, etc. The example in teh results highlights how redundant variables can be identified as a necessary part of the model calibration process.
%' 
%' Why did identifiability decrease for top five parameters in each category after removing redundant phyto/zoop groups?
%' 
%' \citep{Denman03} describes over-parameterization for multiple phyto groups, use this to emphasize collinearity issues with the phyto groups despite differing sensitivity values between groups
%' 
%' Identifiability by category - varies with number of parameters in the category but some were more redundant than others (phytoplankton).  
%' 
%' Questions specific to GOM - what initial conditions are important? How many phytoplankton groups do we need (e.g., related to structural uncertainty)?
%' 
%' How does the assimilation of additional parameters (e.g., other state variables) during calibration influence the conclusions?  \citet{Wagener01} describes this as a potential approach to improving model performance by improving the availability of information for model calibration (p. 14).  
%' 
%' How does uncertainty translate to what a model should provide (generality v precision)?  The first step - find out what can be optimized but then do not overfit....
%' 
%' What about structural uncertainty - does sensitivity of a model to variation in a parameter imply parameter uncertainty and/or structural uncertainty?
%' 
%' A final point about optimization with identifiable parameter sets - optimization to fit the data still does not ensure a correct model.  Failing in one way can be over-compensated by another feature, e.g., the parameter set that is optimized (see \cite{Flynn05}, p. 1207, third paragraph), also \citep{Durand02,Arhonditsis08}, also note that over-parameterized models are not necessarily bad, see \citep{Omlin01}, models must be validated for the uses which they were intended where a comparison of observed and predicted is the simplest approach and other methods should be used depending on analysis needs \citep{Rykiel96}
%' 
%' \cite{Omlin01} state that the sensitivity, identifiability, estimation process is iterative (p. 113), need to rinse and repeat for proper calibration. 
%' 
%' How to improve identifiability - get more/better observed data, include obs from other state variables in RSS minimization (eqn q in \cite{Omlin01})
%' 
%' Alternative methods for uncertainty analysis - bayesian, MCMC, nonlinear calibration-constrained optimization \citep{Gallagher07}, \citep{Arhonditsis08}
%' 
%' Our heuristic products are partially analogous to the Rainfall-Runoff Modelling Toolbox (RRMT) presented by \citep{Wagener01b}2001 (cited on page 15, in \citealt{Wagener01}), although ours is simple in comparison

>>>>

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^2$, Lisa L. Lowe$^3$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Email: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}}}
	  \\\\{\textit {\normalsize $^2$Dauphin Island Sea Lab, University of South Alabama}}
  \\{\textit {\normalsize Dauphin Island, AL 36528}}
	\\{\textit {\normalsize Phone: 251-861-2141 ext. 7567}}
	\\{\textit {\normalsize Email: \href{mailto:jlehrter@disl.org}{jlehrter@disl.org}}}
	\\\\{\textit {\normalsize $^3$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Fri Dec 2 14:10:33 2016 -0600
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

\begin{abstract}
\noindent Bio-geo-chemical models are useful tools in environmental sciences that can guide management and policy-making. Consequently, significant time and resources are spent developing these models in system-specific contexts. The optimization of model parameters to maximize precision, including transferability of these models to different systems, are fundamental concerns in the development and application of these tools. This study provides a context for understanding quantitative limitations of coupled hydrodynamic-ecological models by evaluating parameter sensitivity and identifability of a \ac{zerod} unit of a larger spatio-temporal model of hypoxia on the \acl{lcs} of Gulf of Mexico.  The analysis provides a contrast of numeric and ecological certainty in parameter subsets using a systematic framework to infer larger trends in dissolved oxygen dynamics over time, having implications for understanding factors that contribute to environmental conditions that are detrimental to aquatic resources.  In particular, we focus on issues of parameter identifiability using local sensitivity analyses to provide quantitative descriptions of numerical constraints on model precision.  We argue that quantitative and ecological certainty in model calibration are often at odds and practitioners must choose explicit model components to optimize given tradeoffs between the two. We further conclude that numerically optimal parameter sets for models of hypoxia are often small subsets of the complete parameter set because of redundancies in the unique effects of parameter perturbations on model output.  As a result, we demonstrate that use of a model for inference into ecological mechanisms of observed or predicted changes in hypoxic condition can be potentially misguided in the absence of quantitative descriptions of identifiability.  Although these concerns have been expressed in the literature, they are rarely explicitly addressed or included in model evaluations.  In addition to immediate implications for regional models, we provide a framework for describing the effects of parameter uncertainty and identifiability that can be applied to similar models to better inform environmental management.
\end{abstract}
\acresetall

@@

\noindent where $\gamma$ ranges from one to infinity for perfectly identifiable (orthogonal) or unidentifiable (perfectly collinear) results for parameters in a sensitivity matrix $S$.  The sensitivity functions were supplied as a matrix $\hat{S}$ with rows $i$ and columns $j$ (\cref{sijeqn}) that described deviations of predicted \ac{do} from the default parameter values.  Thus, $\gamma$ can be estimated for any subset of parameter combinations using the change in model output for perturbations of individual parameters. Sensitivity matrices were first normalized by dividing by the square root of the summed residuals \citep{Omlin01,Soetaert10}.

The collinearity index $\gamma$ provides a measure of the linear dependence between sensitivity functions (i.e., $S_i$ for $j$ parameters) described above for subsets of parameters. Estimates of $\gamma$ greater than 10-15 suggest parameter sets are poorly identifiable \citep{Brun01,Omlin01}, meaning parameter values that maximize precision on a calibration dataset are inestimable by conventional optimization algorithms given similar effects of the selected parameters on the estimated state variable. Greater sensitivity of a state variable to a subset of parameters does not always imply better identifiability if the effects of individual parameters are similar.  An intuitive interpretation of $\gamma$ is provided by \citet{Brun01} such that a change in a state variable caused by a change in one parameter can be offset by the fraction $1 - 1/\gamma$ by the remaining parameters.  That is, $\gamma = 10$ suggests the relative change in \ac{do} for an arbitrary parameter in the selected set can be compensated for by 90\% with changes in the other parameters.

Initial analyses suggested that considerably limited subsets of parameters were identifiable of the 51 evaluated for the FishTank model.  Given this limitation, parameter selection must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is the excessively high number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter (e.g., practitioners may only be interested in optics parameters).  For example, \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., based on $n$ choose $k$ combinations equal to $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.  Accordingly, parameter selection is complicated by differing sensitivity, identifiability limits for parameter subsets, and the difficulty of choosing from many combinations.

@@

Local sensitivity analyses for the additional state variables (ammonium, \ac{chla}, irradiance, nitrate, \ac{pom}, \ac{dom}, and phosphorus) had similar results as \ac{do} with some exceptions (\cref{fig:sensalltile,tab:nh4sens,tab:chlsens,tab:irrsens,tab:no3sens,tab:om1sens,tab:om2sens,tab:po4sens}).  All variables were sensitive to the same parameters as \ac{do} (38 of 51 evaluated), although average sensitivity differed between variables.  Average $L1$ ranged from $0.02$ for irradiance (\cref{tab:irrsens}) to $0.71$ for \ac{dom} (\cref{tab:om2sens}).  All average sensitivity values for the state variables were higher than the average for \ac{do} ($L1$ = $0.01$).  For each variable, $L1$ ranged from $2.24\times 10^{-6}$ (\textit{QminP}) to $8.49$ (\textit{mA}) for ammonium (\cref{tab:nh4sens}), $1.38\times 10^{-6}$ (\textit{QminP}) to $13.94$ (\textit{mA}) for \ac{chla} (\cref{tab:chlsens}), $1.92\times 10^{-7}$ (\textit{QminP}) to $0.13$ (\textit{ZKa}) for irradiance (\cref{tab:irrsens}), $6.67\times 10^{-7}$ (\textit{QminP}) to $8.49$ (\textit{umax}) for nitrate (\cref{tab:no3sens}), $6.41\times 10^{-5}$ (\textit{KNH4}) to $7.22$ (\textit{mA}) for \ac{pom} (\cref{tab:om1sens}),  $7.41\times 10^{-5}$ (\textit{KNH4}) to $14.25$ (\textit{mA}) for \ac{dom} (\cref{tab:om2sens}), and $8.21\times 10^{-7}$ (\textit{QminP}) to $1.47$ (\textit{ZKa}) for phosphate (\cref{tab:po4sens}).  For the parameter categories, ammonium was most sensitive to phytoplankton parameters (average $L1$ = $0.8$ across all parameters in the category), \ac{chla} to phytoplankton ($L1$ = $1.14$), irradiance to zooplankton ($L1$ = $0.03$), nitrate to zooplankton ($L1$ = $1.06$), \ac{pom} to temperature ($L1$ = $0.86$), \ac{dom} to temperature ($L1$ = $1.48$), and phosphate to zooplankton ($L1$ = $0.31$).  Finally, average sensitivity between parameter categories independent of the state variables ranged from $0.01$ for optics (average $L1$ across all variables) to $0.62$ for phytoplankton.

@@

The identifiability analyses suggested that many parameter subsets exceeded the thresholds of $\gamma = 10, 15$, providing further justification for using selection heuristics for parameter optimization.  Results for \ac{do} are provided first to demonstrate general concepts for the identifiability analyses, followed by an extension to the remaining state variables.  Parameter identifiability (as $\gamma$) for \ac{do} increased at different rates with increasing size of parameter subsets depending on the parameter category or the number of top parameters that were selected (\cref{fig:identplo}). By category, identifiability was lowest for all combinations of parameter subsets in the phytoplankton ($60$\% of subsets less than $\gamma = 15$, $43$\% less than $\gamma = 10$) and zooplankton categories ($53.1$\% less than $\gamma = 15$, $40$\% less than $\gamma = 10$), whereas all combinations were identifiable for optics ($100$\% less than $\gamma = 15$, $100$\% less than $\gamma = 10$) and a majority identifiable for organic matter ($91.9$\% less than $\gamma = 15$, $76.5$\% less than $\gamma = 10$). Identifiability for parameters in the temperature category was not evaluated because \ac{do} was sensitive to only one parameter (i.e., $\gamma = 1$). Parameter combinations for choosing from the top, top two, top three, and top four parameters in each category together had decreasing identifability with the increasing size of the selection pool (e.g., top one versus top four parameters, \cref{fig:identplo}).  The percentage of parameter subsets that were below the acceptable thresholds for identifability was $100$\% less than $\gamma = 15$ and $100$\% less than $\gamma = 10$ for the top parameters in each category, $90.6$\% and $80.7$\% for the top two, $80.7$\% and $70.9$\% for the top three, and $55.8$\% and $45.7$\% for the fop four.  Results for the remaining state variables had similar patterns in identifiability with increasing size of parameter subsets and selection categories, although differences in identifiabiltiy between state variables was observed (\cref{fig:identploall}).  Most notably, nitrate was consistently the least identifiable variable for all selection heuristics (highest $\gamma$), whereas \ac{do} was most identifiable.

An alternative view of the results in \cref{fig:identplo} can be used to demonstrate the effects of parameter selection criteria and number of parameters in the selection pool on identifiability.  \Cref{fig:percthresh} shows the percentage of identifiable parameter sets for \ac{do} using the same selection criteria in \cref{fig:identplo}, i.e., selection of parameters only within parameter categories and selection of the top sensitive parameters regardless of category.  \Cref{fig:percthresh} is conceptually similar to \cref{fig:identplo}, with the added effect of a chosen $\gamma$ threshold on identifability.  Previous studies have provided general rules for $\gamma$ thresholds \citep{Brun01,Omlin01}, such that exact values beyond which model calibration fails could vary for the given model and optimization method.  The selected $\gamma$ thresholds are somewhat arbitrary, although the results show differences that have implications for parameter selection for model calibration.  In general, identifiability decreased with the addition of parameters, although the rate of decrease depended on the selected threshold for $\gamma$. More conservative values for $\gamma$ (e.g., $\gamma = 5$) were more sensitive to the number of parameters in a subset, that is, identifiability decreased more quickly with the addition of parameters at lower $\gamma$ thresholds as compared to higher $\gamma$.  Notable differences in identifiability were also observed by selection criteria (within categories or top parameters only), which further supports results in \cref{fig:identplo,fig:identploall}.

An evaluation of the effects of individual parameters on $\gamma$ suggested that some parameters had disproportionate effects on identifiability.  Based on $\gamma = 15$, \cref{fig:identplo} suggests that most parameter sets for organic matter were identifiable, regardless of how many parameters were selected (i.e., two through eight).  However, some subsets were not identifiable such that identification of one or more redundant parameters that are inflating $\gamma$ values could provide useful information.  \Cref{fig:exclex} shows an alternative view of identifiability of \ac{do} with exclusion and inclusion of individual parameters in different sets for the organic matter category.  As before, collinearity increases with more parameters in a subset, although the increase varies depending on which parameter was included or excluded from the set.  For example, inclusion of \textit{KNO3} in a parameter set almost always inflated $\gamma$.  All parameter subsets that did not include \textit{KNO3} were well below $\gamma = 15$, suggesting that removal of this parameter improves identifiability.  Interestingly, the inclusion of some parameters caused a reduction in $\gamma$, which contradicts the general rule that more parameters caused reduced identifiablity.  For example, parameter sets that included \textit{KGcdom} generally had lower $\gamma$ values relative to those that excluded the parameter.

@@

The above results demonstrated that the state variables differed in the magnitudes of sensitivity for each parameter and the number of identifiable subsets, where the latter varied by identifiability thresholds and parameter selection criteria.  These results further underscore the need for explicit selection heuristics for parameters to calibrate for improving model performance. Results for each of the three selection heuristics (blocked by parameter category, independent of category, all categories equally) applied to each state variable differed in the number of selected parameters and distribution of parameters within each category (\cref{tab:heurist1,tab:heurist2,tab:heurist3}).  In general, a corresponendence was observed between the number of parameters that were selected given the threhold of $\gamma = 15$ and relative identifiability between the state variables.  As noted above, nitrate was the least identifiable variable (\cref{fig:identploall}), whereas other variables (e.g., \ac{do}, irradiance) were more identifiable.  The constraints on relative identifiability between variables were demonstrated with the selection heuristics.  For example, heuristics for nitrate typically selected only one or two parameters that met the criteria as compared to more identifiable variables that included several parameters that were identifiable. Overall, the first selection heuristic demonstrated that the number of parameters chosen by parameter category differed independently of the state variables (\cref{tab:heurist1}). The number of selected parameters averaged across state variables in decreasing order was 4.25 parameters from the phytoplankton category, 3.5 from organic matter, 2.75 from optics, and 2.38 from zooplankton.  These rankings were independent of parameter sensitivity and identifiability in each category.  The second and third selection heuristics (\cref{tab:heurist2,tab:heurist3}) were similar, although more parameters were generally selected for the third heuristic.

\Cref{fig:heurist_stts} demonstrates parameter selection for all state variables following the second heuristic where parameters were chosen by decreasing sensitivity independent of parameter categories (exact values are shown in \cref{tab:heurist2}).  The y-axis on the plot shows the relative identifiability values with the addition of parameters from one to many on the x-axis (from left to right).  The second to last parameter for each variable is the last parameter selected within the potential threshold of $\gamma = 15$.  Interestingly, the last parameter shown for most of the state variables caused a relatively large increase in $\gamma$ that was disproportionate to the combined identifiability of the preceding parameter set. For most variables, the phytoplankton edibility vector for zooplankton (\textit{ediblevector(Z1)}) caused a dramatic increase in $\gamma$ with inclusion in the parameter set.  In addition to demonstrating the approach for selection with the second heuristic, \cref{fig:heurist_stts} provides an alternative method of identifying parameters that are disporportionately redundant within a given parameter set.  This infomation is useful if additional parameter selection rules are developed independently of those proposed herein.

@@

\section{Discussion}

Common goals in the application of biogeochemical models of ecosystem processes are to \begin{inparaenum}[1\upshape)]
\item accurately describe the system with historical data, and 
\item provide a means of forecasting ecosystem condition with hypothesized management or environmental scenarios. 
\end{inparaenum}
Although these objectives are the focus of most model applications, the structural components of process-based models should secondarily provide inference into which ecosystem processes and functions are driving observed or hypothesized changes.  This latter objective represents a more general scientific outcome of biogeochemical models that extends beyond the applied benefits of describing and predicting change in a particular system.  Modelers often hope to identify universal principles that govern dynamics across systems and the constraint of model parameters to observations provides a means of supporting or refuting these principles.  Extension of these principles to test the effects of structural changes and observation uncertainty on model predictions provideds further evidence to validate model components. This study provided a simple approach to use the effects of parameter perturbations on model state variables to characterize identifiable parameter subsets that vary by parameter selection criteria.  By doing so, we demonstrated that small parameter subsets relative to all sensitive parameters were within the identifiabilty thresholds described in the literature.  The identifiable parameter subsets varied considerably between state variables and the method for parameter selection.  We further demonstrated that changes in the model structure and variation in the initial conditions had an effect on sensitivity which has direct implications for identifiability. In general, these results provide justification for the use of explicit parameter selection heuristics that practitioners should adopt for model calibration.

Although the results were specific to individual variables, some generalities can be inferred from the sensitivity analsyes. State variables were most sensitive to parameters in the phytoplankton and zooplankton categories, particularly the maximum growth rates (\textit{umax} for phytoplankton, \textit{Zumax} for zooplankton), mortality coefficient for phytoplankton (\textit{mA}), and the zooplankton half saturation coefficent for grazing (\textit{ZKa}). An increase in growth rate of primary producers has the potential to increase oxygen concentration through photosynthetic processes, although increased production of organic matter is balanced with respiration and bacterial decomposition that reduce \ac{do} concentration in the water column.  Similarly, increases in zooplankton abundance with increased growth rates causes a reduction in phytoplankton biomass through grazing, which is expected to further deplete pool of organic matter.  Most variables were also sensitive to variation in the half-saturation grazing coefficient which is linked to the concentration of nutrients that support half the maximum grazing rate. Although the tradeoff between abundance, grazing, and decomposition is complex, the sensitivity of model state variables to parameters that directly control the abundance of primary producers is in agreement with empirical observations of factors that influence hypoxia dynamics on the \ac{lcs}. The sensitivity of the model output to variation in other parameters that relate to physical and chemical properties of the system was of secondary importance to biological relationships.  State variables were sensitive to changes in light and temperature parameters, although to a lesser extent than phytoplankton and zooplankton parameters.  As such, the differing sensitivities of state variables to parameters in each of the categories was not unexpected given general ecological relationships that are well understood and described by the model.
